RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-10-10 discussion.tex
Related work & references cleanup

The need for justifications of modeling choices plays an important role in different requirements engineering methods using goal models. The belief element in the original GRL specification \cite{Amyot:2010:EGM:1841349.1841356} is meant to capture the rationales behind the inclusion of goals and tasks in the model. Furthermore, relations between elements in a goal model also provide justifications: high-level goals are reasons for lower-level goals, tasks and resources. Hence, refinement and decomposition techniques used in requirements engineering \cite{van2001goal} can be seen as explicit argumentation steps in the goal modeling process. Take, for example, CQ2 of PRAS (Section~\ref{sect:background:pras}), which asks whether there are alternative ways of realizing the same goal. Providing an alternative sub-goal or -task in a goal model then an explicit argumentative move in the discussion.

So in a sense, a goal model provides a justification for itself, particularly if we include belief elements for extra design rationalization. This idea is also prevalent in our RationalGRL framework: many arguments are in fact GRL elements, and many critical questions can be answered by introducing (\textsf{INTRO}) new GRL elements. However, as was already argued in Section~\ref{sect:introduction} (and also by \cite{Jureta:RE2008}), argumentation produces different, richer and complementary information to just the goal model. The goal model is the product of a process of argumentation, and does not include, for example, goals and tasks that were at some point considered discarded. Furthermore, for goal models it is only possible to determine the satisfiability of goals given the possible tasks and resources; what cannot be determined is the acceptability of goals, that is, whether they are acceptable given potentially contradictory opinions of stakeholders.

Closely related to goal modeling and design justification are frameworks that aim to provide a design rationale (DR)~\cite{shum2006hypermedia}, an explicit documentation of the reasons behind decisions made when designing a system or artefact. DR looks at issues, options and arguments for and against these options in the design of, for example, a software system. The similarity with goal modeling is that much of the traditional DR literature provides modeling languages and tool support for building DR diagrams. The differences lie in the concepts used: in DR, the focus is more on the rationales, why certain design options were chosen, than on the functional and non-functional requirements of the system, which are expressed in goal models. The RationalGRL framework in a sense incorporates the ideas from DR into goal models by including arguments and issues (critical questions) into the RatinalGRL model.

Arguments and requirements engineering approaches have been combined by, among others, Haley \emph{et al.}~\cite{haley2005arguing}, who use structured arguments to capture and validate the rationales for security requirements. However, they do not use goal models, and thus, there is no explicit trace from arguments to goals and tasks. Furthermore, like~\cite{Jureta:RE2008}, the argumentative part of their work does not include formal semantics for determining the acceptability of arguments, and the proposed frameworks are not actually implemented. Murukannaiah \emph{et al.}~\cite{murukannaiah2015resolving} propose Arg-ACH, an approach to capture inconsistencies between stakeholders' beliefs and goals, and resolve goal conflicts using argumentation techniques.

@@

\todo{Marc}{all}{Review the following notes:

\begin{itemize}
\item RE17 paper: merging argumentation theory with law
  "Using Argumentation to Explain Ambiguity in Requirements Elicitation Interviews" Yehia Elrakaiby, Alessio Ferrari, Paola Spoletini, Stefania Gnesi and Bashar Nuseibeh
\item more details on Murukannaiah [RE15]
\item related work on argumentation theory - check Serena Villata on goal models
\item Prakken and Wierenga
\item First paper we wrote they also said something about Bashar Nuseibeh
\item Also mention argumentation in AI - arg diagramming - and how that relates to our work
\item Software design discussions Black et al.
\item Maybe this one:
  "Complete Traceability for Requirements in Satisfaction Arguments"
  Anitha Murugesan, Michael Whalen, Elaheh Ghassabani and Mats Heimdahl
  \end{itemize}
}
\subsection{Open issues}
\label{sect:goalmodeling:openissues}

>>>>

\paragraph{Design Rationale} Argumentation in software design has for some time now been the subject of the work on so-called \emph{design rationale} (DR)~\cite{shum2006hypermedia}, an explicit documentation of the reasons behind decisions made when designing a system or software architecture. DR looks at issues, options and arguments for and against these options in the design of, for example, a software system. Similar to the literature on goal modeling, much of the traditional DR literature provides modeling languages and diagramming tool support for building design rationales. It is in this diagramming functionality that the link with argument diagrams from philosophy, law and AI~\cite{reed2004araucaria,gordon2007visualizing} has been made, where argument diagrams represent reasoning from premises to conclusions. More recent work on DR moves away from the idea that all decision have to be explicitly diagrammed and focuses more on empirically investigating how critical reflection can help when designing \cite{razavian2016two,SchriekEtal2016}, or which parts of the design process are best explicitly documented \cite{falessi2013value}.

Software design and requirements engineering are very closely related \cite{nuseibeh2001weaving} and hence the insights from the DR literature are directly applicable to RE. The work on the RationalGRL framework essentially incorporates the core ideas from DR into goal-oriented requirements engineering by explicitly including arguments pro and con the various options into the goal model, and by proposing a methodology and critical questions that encourage reflection when thinking about the possible goals and functionality of a system.

\paragraph{Requirements Engineering} There are a number of general approaches in the field of requirements engineering that explicitly take into account arguments. One early example comes from Haley et al.~\cite{haley2008security}, who use formal logical arguments to show that the system behavior satisfies certain security requirements, and more informal arguments to capture and validate the assumptions underlying the system behavior. This system behavior is defined by the tasks it executes and thus arguments are given for and against system tasks, similar to the way beliefs and counterarguments can be provided for tasks in RationalGRL. What Haley et al. leave implicit in their argumentation are the goals of the stakeholders on which the system tasks depend -- they include the goals in their framework and mention that there will often be conflicting goals between stakeholders, but do not explicitly model them. Furthermore, the argumentative part of their framework does not include formal semantics for resolving conflict between arguments or determining the acceptability of arguments. Yu et al.~\cite{yu2015automated} further extend the framework by Haley et al., including algorithms for Dung-style \cite{Dung1995} argumentation semantics and a database of specific ways in which to attack (or mitigate) risks, which can be likened to a set of critical questions for risks and security requirements (cf. Yu et al.~\cite{yu2015automated} Section 3.1).

Another recent example of the use of arguments in goal-oriented requirements engineering is the work by Murukannaiah et al.~\cite{murukannaiah2015}, who propose Arg-ACH, an approach 
in which the beliefs underlying conflicting goals can be made explicit using argumentation. Murukannaiah et al. start with the basic technique of Analysis of Competing Hypotheses (ACH), where for conflicting goals the beliefs that are consistent and inconsistent with these goals are included in a matrix and counted. They then extend this technique into Arg-ACH: instead of just indicating whether a belief is consistent or inconsistent with a goal, each belief becomes an argument for or against the goal, which is then diagrammed using the Carneades tool \cite{gordon2007visualizing}. Belief scores are assigned to arguments, which can be aggregated to provide one's belief in a goal. The arguments for and against goals can be based on argument schemes, and critical questions can be used to find new arguments for or against the goals or the existing arguments. One example provided in the paper is the argument scheme from expert opinion, which allows one to draw conclusions based on expert statements and subsequently question, for example, the objectivity and veracity of the expert using critical questions. Murukannaiah et al. conducted an experiment in which they had two groups, one with ACH and one with Arg-ACH, perform an analysis of several conflicting goals regarding security at transport hubs. They found that, while the group that used Arg-ACH took longer, they also covered more possibilities in their belief search and the conclusions were more consistent among the group.

One other example of argumentation in RE concerns the use of argumentation in requirements elicitation. Ionita et al.~\cite{ionita2014argumentation} propose an simple argumentation dialogue game in which risk assessors try to attack each other's arguments for certain risks attached to a system design. Dung's semantics \cite{Dung1995} are then used to determine the risks that are still valid, and those that have been successfully rejected. Elrakaiby et al.~\cite{ElrakaibyFSGN17} use argumentation to explain ambiguity. They define when a statement by a client who is being interviewed about the requirements of a system presents an inconsistency (either with the client's previous statements or with the requirement engineer's beliefs) or an insufficiency, that is, when an analyst needs more information from the client to accept a client's statement. The inconsistencies are then captured as mutually attacking arguments, and the insufficiencies as arguments against the original statements saying that, for example, the functionality expressed in the statement cannot be realised or is irrelevant. Elrakaiby et al. coded the data from 34 requirement elicitation interviews, identifying 39 inconsistencies (i.e. at least two arguments that mutually attack) and 29 insufficiencies (i.e., at least one argument attacking another).

It is clear from this literature that arguments play a core role in RE. Murukannaiah et al.~\cite{murukannaiah2015} show that critical reflection using argument schemes and critical questions -- in the same way that RationalGRL proposes -- improves the quality of the reasoning in the RE process. Elrakaiby et al.~\cite{ElrakaibyFSGN17} provide a case study similar to the current one, identifying, as we did, many counterarguments specifically with respect to realisability, relevance and clarity (cf. CQ2a, CQ3, CQ11 and CQ12 in Table~\ref{table:argument-schemes}). Like in RationalGRL, the use of Dung-style argumentation semantics to compute the acceptable claims in RE is further also advocated by the literature \cite{yu2015automated,ionita2014argumentation,ElrakaibyFSGN17}.

The current work on RationalGRL puts the insights from the above-mentioned literature in a broader framework. For example, there is a specific focus on security requirements \cite{haley2008security,yu2015automated,ionita2014argumentation} or the reasoning is about single goals or tasks instead of about the wider context as represented in a goal model \cite{haley2008security,yu2015automated,murukannaiah2015,ElrakaibyFSGN17}. RationalGRL provides a generic and extensible framework for arguing about goals and tasks in RE. At the moment, there is only a ``generic argument'' in addition to arguments about goals and tasks. However, new argument schemes and critical questions about for example, security risks or expert opinions, can be easily added: the metamodel (Figure~\ref{fig:metamodel}) accommodates this and the formal specification in Section~\ref{sect:formalframework} is set up in such a way that  extending the definition of argument and adding new algorithms for specific critical questions is easy.

\paragraph{Goal Modeling} Argumentation has been included -- both explicitly and implicitly -- in existing goal modeling languages. For example, the belief element in the original GRL specification \cite{Amyot:2010:EGM:1841349.1841356} is meant to capture the rationales (reasons, arguments) behind the inclusion of goals and tasks in the model. Furthermore, relations between elements in a goal model also provide justifications: high-level goals are reasons for lower-level goals, tasks and resources. Hence, refinement and decomposition techniques used in requirements engineering \cite{van2001goal} can be seen as explicit argumentation steps in the goal modeling process. Take, for example, CQ2 of PRAS (Section~\ref{sect:background:pras}), which asks whether there are alternative ways of realizing the same goal. Providing an alternative sub-goal or -task in a goal model then an explicit argumentative move in the discussion. So in this sense, a goal model provides a justification for itself, particularly if we include belief elements for extra design rationalization. This idea is also prevalent in our RationalGRL framework: many arguments are in fact GRL elements, and many critical questions can be answered by introducing new GRL elements. However, as was already argued in Section~\ref{sect:introduction} (and also by \cite{Jureta:RE2008}), argumentation produces different, richer and complementary information to just the goal model. The goal model is the product of a process of argumentation, and does not include, for example, goals and tasks that were at some point considered but discarded. Furthermore, for goal models it is only possible to determine the satisfiability of goals given the possible tasks and resources; what cannot be determined is the acceptability of goals, that is, whether they are acceptable given potentially contradictory opinions of stakeholders.

@@

One idea is to capture requirements engineering and software design processes as explicit dialogues between parties~\cite{finkelstein1989multiparty}.

\todo{F}{F}{Review the following:

\begin{itemize}
\item Mirbel and Villata on Dung \& goal models
\item Software design discussions Black et al., Prakken and Wierenga, \cite{finkelstein1989multiparty}
  \end{itemize}
}
\subsection{Open issues}
\label{sect:goalmodeling:openissues}

### 2017-10-10 introduction.tex
Related work & references cleanup

Among the initial activities in RE are the ``early-phase'' requirements engineering activities, which include those that consider how the intended system should meet organizational goals, why it is needed, what alternatives may exist, what the implications of the alternatives are for different stakeholders, and how the interests and concerns of stakeholders might be addressed~\cite{yu1997towards}. These activities fall under the umbrella of goal modeling. There are a large number of established RE methods using goal models in the early stage of requirements analysis (e.g.,~\cite{liu2004designing,donzelli2004goal,dardenne1993goal,chung2012non,castro2002towards}, overviews can be found in~\cite{van2001goal,kavakliL05}). Several goal modeling languages have been developed in the last two decades as well. The most popular ones include $i*$~\cite{Yu:1997:TMR:827255.827807}, Keep All Objects Satisfied (KAOS)~\cite{van2008requirements}, the NFR framework~\cite{chung2012non}, \textsc{Tropos}~\cite{giorgini2005goal}, the Business Intelligence Model (BIM)~\cite{horkoff2014strategic}, and the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}.

>>>>

Among the initial activities in RE are the ``early-phase'' requirements engineering activities, which include those that consider how the intended system should meet organizational goals, why it is needed, what alternatives may exist, what the implications of the alternatives are for different stakeholders, and how the interests and concerns of stakeholders might be addressed~\cite{yu1997towards}. These activities fall under the umbrella of goal modeling. There are a large number of established RE methods using goal models in the early stage of requirements analysis (overviews can be found in~\cite{kavakliL05,van2001goal}). Several goal modeling languages have been developed in the last two decades as well. The most popular ones include $i*$~\cite{yu1997towards}, Keep All Objects Satisfied (KAOS)~\cite{van2008requirements}, the NFR framework~\cite{chung2012non}, \textsc{Tropos}~\cite{giorgini2005goal}, the Business Intelligence Model (BIM)~\cite{horkoff2014strategic}, and the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}.

### 2017-10-10 main.tex
Section 1-4
Small changes 1, 2, some work on case study section (3); section 4 is
not finished yet!
Finished section 4 (for now)
changes up to formal framework, publisher files
notes formal framework and small changes
Related work & references cleanup

\begin{document}
\twocolumn[{%
 \centering
 \LARGE RationalGRL: A Framework for Argumentation and Goal Modeling \\[1.5em]
 \large Marc van Zee\Mark{1},
        Floris Bex\Mark{2},
        and Sepideh Ghanavati\Mark{3}\\[1em]
 \normalsize
 \begin{tabular}{*{3}{>{\centering}p{.33\textwidth}}}
  \Mark{1}Computer Science and Communication (CSC) & \Mark{2}Department of Information and Computing Sciences & \Mark{3}Department of Computer Science \tabularnewline
  University of Luxembourg & Utrecht University & Texas Tech University\tabularnewline
  \url{marcvanzee@gmail.com} & \url{f.j.bex@uu.nl} & \url{sepideh.ghanavati@ttu.edu}
 \end{tabular}\\[3em] % some more space after the title part
}]

@@

\input{introduction}
\input{background}
\input{overview}
\input{gmas}
\input{formalframework}
\input{tool}
\input{discussion}

>>>>

\begin{document}
\twocolumn[{%
 \centering
 \LARGE RationalGRL: A Framework for Argumentation and Goal Modeling \\[1.5em]
 \large Marc van Zee\Mark{1},
        Floris Bex\Mark{2}
        and Sepideh Ghanavati\Mark{3}\\[1em]
 \normalsize
 \begin{tabular}{*{3}{>{\centering}p{.33\textwidth}}}
  \Mark{1}Computer Science and Communication (CSC) & \Mark{2}Department of Information and Computing Sciences & \Mark{3}Department of Computer Science  \tabularnewline
  University of Luxembourg & Utrecht University & Texas Tech University \tabularnewline
  \url{marcvanzee@gmail.com} & \url{f.j.bex@uu.nl} & \url{sepideh.ghanavati@ttu.edu}
 \end{tabular}\\[3em] % some more space after the title part
}]

@@

\input{introduction}
\input{background}
\input{gmas}
\input{overview}
%\input{examples}
\input{tool}
\input{formalframework}
\input{discussion}
